---
title: "Final Project: The Evolution of Heat in Hot Ones"
author: "Balazs Gyarmati"
date: "2025-12-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(tidytuesdayR)
library(broom)
```

#In this project, I analyze the dataset from the popular YouTube show "Hot Ones", where celebrities answer questions while eating increasingly spicy chicken wings. The data comes from the TidyTuesday archive (2023-08-08).
#My specific research question is: Has the show actually become "hotter" over the years? I will investigate if there is a statistically significant increase in the Scoville heat units of the sauces used across the seasons.

```{r Data reading}
tuesdata <- tidytuesdayR::tt_load('2023-08-08')
sauces <- tuesdata$sauces
```

```{r Exploratory Data Analysis}
#First, I want to understand the distribution of the sauce heat. Since Scoville scores can range from zero to millions, I expect a skewed distribution.

#Histogram of Scoville scores
sauces %>% 
  ggplot(aes(x = scoville)) +
  geom_histogram(fill = "firebrick", color = "white", bins = 30) +
  scale_x_log10(labels = scales::comma) + # Log scale makes the outliers readable
  theme_minimal() +
  labs(title = "Distribution of Sauce Heat (Scoville)",
       subtitle = "Most sauces are mild, but some push the scale to extreme limits",
       x = "Scoville Score (Log Scale)",
       y = "Count")

#Next, I visualize the trend. I will calculate the average heat per season and plot it to see the progression.

#Calculating average heat per season
avg_heat_per_season <- sauces %>% 
  group_by(season) %>% 
  summarise(avg_scoville = mean(scoville, na.rm = TRUE))

#Plotting the trend
avg_heat_per_season %>% 
  ggplot(aes(x = season, y = avg_scoville)) +
  geom_line(color = "orange", size = 1.2) +
  geom_point(color = "red", size = 3) +
  geom_smooth(method = "lm", color = "darkred", linetype = "dashed") +
  theme_light() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Average Sauce Heat per Season",
       subtitle = "There is a clear upward trend in the average spiciness",
       x = "Season Number",
       y = "Average Scoville Score")
```

```{r Statistical Modeling}
#My hypothesis is that the Season number has a positive and significant effect on the Scoville score.

model_avg <- lm(avg_scoville ~ season, data = avg_heat_per_season)
tidy(model_avg)

#Interpretation:
#Estimate (Slope): The coefficient for season is positive and statistically significant. It suggests that with every new season, the average heat of the lineup increases by 7453.091 Scoville units.
#P-value: The p-value is extremely small (less than 0.003), which confirms that this upward trend is statistically significant. We can reject the null hypothesis; the correlation between seasons and spice level is not due to random chance.
```

```{r Residual diagnostics, assumption checks}
#To ensure these results are valid, I check the diagnostic plots of our new linear model.
par(mfrow = c(2, 2))
plot(model_avg)

#Residuals vs Fitted: The red line is relatively horizontal, which indicates that a linear model is a good fit for this data. There is no strong non-linear pattern (like a U-shape) that we missed.
#Normal Q-Q: The points follow the diagonal dotted line much better than in the raw data model. This indicates that the residuals are normally distributed, satisfying a key assumption of linear regression.
```

#Conclusion
#In this project, I analyzed the evolution of sauce spiciness in the show Hot Ones.
#Initial exploration showed massive variance within seasons, but by aggregating the data, a clear pattern emerged. My statistical model confirms a significant positive relationship between the season number and the average Scoville rating.
#The data supports the conclusion that the showrunners have consistently "turned up the heat" over the years, making the challenge progressively more difficult for guests in later seasons.